---
title: "Biostat 203B Homework 5"
subtitle: Due Mar 20 @ 11:59PM
author: "Tanya Wang, 605587605"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

## Predicting ICU duration

Using the ICU cohort `mimiciv_icu_cohort.rds` you built in Homework 4, develop at least three machine learning approaches (logistic regression with enet regularization, random forest, boosting, SVM, MLP, etc) plus a model stacking approach for predicting whether a patient's ICU stay will be longer than 2 days. You should use the `los_long` variable as the outcome. You algorithms can use patient demographic information (gender, age at ICU `intime`, marital status, race), ICU admission information (first care unit), the last lab measurements before the ICU stay, and first vital measurements during ICU stay as features. You are welcome to use any feature engineering techniques you think are appropriate; but make sure to not use features that are not available at an ICU stay's `intime`. For instance, `last_careunit` cannot be used in your algorithms. 


### Question 1
Data preprocessing and feature engineering.


**Solution:**

Load the dataset and convert categorical variables to factor:
```{r}
library(tidyverse)
library(caret)
library(data.table)

mimic_icu_cohort <- 
  readRDS("~/203b-hw/hw4/mimiciv_shiny/mimic_icu_cohort.rds") |>
  filter(!is.na(los_long)) |>
  select(-los, -last_careunit, -dod, -discharge_location, -hospital_expire_flag,
         -intime, -outtime, -admittime, -dischtime, -deathtime, 
         -edregtime, -edouttime, -admit_provider_id, -anchor_year, -anchor_age, 
         -anchor_year_group) |>
  mutate(across(
    c(los_long,
      insurance, 
      language, 
      marital_status, 
      gender), as.factor))

colnames(mimic_icu_cohort)
```




```{r}
str(mimic_icu_cohort)
```


```{r}
library(ggplot2)
library(gridExtra)

# Select only continuous variables
continuous_vars <- mimic_icu_cohort |>
  select(
    bicarbonate, chloride, creatinine, glucose, potassium, sodium, 
    hematocrit, wbc, heart_rate, 
    non_invasive_blood_pressure_diastolic, 
    non_invasive_blood_pressure_systolic, 
    respiratory_rate, temperature_fahrenheit, age_intime
  )

plot_list <- list()

for (var in colnames(continuous_vars)) {
  p <- ggplot(mimic_icu_cohort, aes_string(x = var)) +
    geom_density(fill = "blue", alpha = 0.5, na.rm = TRUE) +
    labs(title = var, x = "Value", y = "Density") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.1, face = "bold", size = 8))
  
  plot_list[[var]] <- p
}

grid.arrange(grobs = plot_list, ncol = 4)
```

Handle missing values (for continuous variables, impute missing values using the median, and
for categorical variables, replace missing values with the mode):
```{r}
impute_missing <- function(df) {
  for (col in colnames(df)) {
    if (is.numeric(df[[col]])) {
      df[[col]][is.na(df[[col]])] <- median(df[[col]], na.rm = TRUE)
    } else {
      df[[col]][is.na(df[[col]])] <- 
        as.character(names(sort(table(df[[col]]), decreasing = TRUE))[1])
    }
  }
  return(df)
}

mimic_icu_cohort <- impute_missing(mimic_icu_cohort)
colSums(is.na(mimic_icu_cohort))
```



### Quetion 2
Partition data into 50% training set and 50% test set. Stratify partitioning according to `los_long`. For grading purpose, sort the data by `subject_id`, `hadm_id`, and `stay_id` and use the seed `203` for the initial data split. Below is the sample code.
```{r}
#| eval: false
set.seed(203)

# sort
mimiciv_icu_cohort <- mimiciv_icu_cohort |>
  arrange(subject_id, hadm_id, stay_id)

data_split <- initial_split(
  mimiciv_icu_cohort, 
  # stratify by los_long
  strata = "los_long", 
  prop = 0.5
  )
```


**Solution:**
```{r}
library(tidymodels)

set.seed(203)

# Sort dataset by subject_id, hadm_id, stay_id
mimic_icu_cohort <- mimic_icu_cohort |>
  arrange(subject_id, hadm_id, stay_id)

# Split the dataset
data_split <- initial_split(
  mimic_icu_cohort, 
  strata = "los_long",
  prop = 0.5
)

train_data <- training(data_split) |>
  select(-subject_id, -hadm_id, -stay_id)

test_data <- testing(data_split) |>
  select(-subject_id, -hadm_id, -stay_id)

dim(train_data)
dim(test_data)
```


### Question 3
Train and tune the models using the training set.

#### Logistic regression (with enet regularization)

```{r}
logit_recipe <- recipe(
  los_long ~ ., 
  data = train_data
  ) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  print()
```

```{r}
logit_mod <- 
  logistic_reg(
    penalty = tune(), 
    mixture = tune()
  ) |> 
  set_engine("glmnet", standardize = FALSE) |>
  print()
```

```{r}
logit_wf <- workflow() |>
  add_recipe(logit_recipe) |>
  add_model(logit_mod) |>
  print()
```

```{r}
param_grid <- grid_regular(
  penalty(range = c(-6, 3)), 
  mixture(),
  levels = c(5, 5)
  ) |>
  print()
```

Set cross-validation partitions.
```{r}
set.seed(203)

folds <- vfold_cv(train_data, v = 5)
folds
```

Fit cross-validation.
```{r}
library(glmnet)

(logit_fit <- logit_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )) |>
  system.time()

logit_fit
```

Visualize CV results:
```{r}
logit_fit |>
  # aggregate metrics from K folds
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = penalty, y = mean, color = factor(mixture))) +
  geom_point() +
  labs(x = "Penalty", y = "CV AUC") +
  scale_x_log10()
```

Show the top 5 models.
```{r}
logit_fit |>
  show_best(metric = "roc_auc")
```

Select the best model.
```{r}
best_logit <- logit_fit |>
  select_best(metric = "roc_auc")
best_logit
```

Fit this final model to the whole training data and use test data to estimate the model performance.
```{r}
# Final workflow
final_wf <- logit_wf |>
  finalize_workflow(best_logit)
final_wf
```

```{r}
# Fit the whole training set, then predict the test cases
final_fit <- 
  final_wf |>
  last_fit(data_split)
final_fit
```

```{r}
# Test metrics
final_fit |> 
  collect_metrics()
```


#### Random Forest

```{r}
rf_recipe <- 
  recipe(
    los_long ~ ., 
    data = train_data
  ) |>
  step_zv(all_numeric_predictors())
```

```{r}
rf_mod <- 
  rand_forest(
    mode = "classification",
    # Number of predictors randomly sampled in each split
    mtry = tune(),
    # Number of trees in ensemble
    trees = tune()
  ) |> 
  set_engine("ranger")
rf_mod
```

```{r}
rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_mod)
rf_wf
```

```{r}
param_grid <- grid_regular(
  trees(range = c(100L, 500L)), 
  mtry(range = c(1L, 5L)),
  levels = c(5, 5)
  )
param_grid
```

Set cross-validation partitions.
```{r}
set.seed(203)

folds <- vfold_cv(train_data, v = 5)
folds
```

Fit cross-validation.
```{r}
library(ranger)

rf_fit <- rf_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
rf_fit
```

Visualize CV results:
```{r}
rf_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = trees, y = mean, color = factor(mtry))) +
  geom_point() + 
  # geom_line() + 
  labs(x = "Num. of Trees", y = "CV AUC")
```

Show the top 5 models.
```{r}
rf_fit |>
  show_best(metric = "roc_auc")
```

Select the best model.
```{r}
best_rf <- rf_fit |>
  select_best(metric = "roc_auc")
best_rf
```

```{r}
# Final workflow
final_wf <- rf_wf |>
  finalize_workflow(best_rf)
final_wf
```

```{r}
# Fit the whole training set, then predict the test cases
final_fit <- 
  final_wf |>
  last_fit(data_split)
final_fit
```

```{r}
# Test metrics
final_fit |> 
  collect_metrics()
```


#### XGBoost

```{r}
library(xgboost)

gb_recipe <- 
  recipe(
    los_long ~ ., 
    data = train_data
  ) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_numeric_predictors()) |> 
  print()
```

```{r}
gb_mod <- 
  boost_tree(
    mode = "classification",
    trees = 500, 
    tree_depth = tune(),
    learn_rate = tune()
  ) |> 
  set_engine("xgboost")
gb_mod
```

```{r}
gb_wf <- workflow() |>
  add_recipe(gb_recipe) |>
  add_model(gb_mod)
gb_wf
```

```{r}
param_grid <- grid_regular(
  tree_depth(range = c(3L, 10L)),
  learn_rate(range = c(-3, 2), trans = log10_trans()),
  levels = c(3, 10)
  )
param_grid
```

Set cross-validation partitions.
```{r}
set.seed(203)

folds <- vfold_cv(train_data, v = 5)
folds
```

Fit cross-validation.
```{r}
gb_fit <- gb_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
gb_fit
```

Visualize CV results:
```{r}
gb_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = learn_rate, y = mean, color = factor(tree_depth))) +
  geom_point() +
  labs(x = "Learning Rate", y = "CV AUC") +
  scale_x_log10()
```

Show the top 5 models.
```{r}
gb_fit |>
  show_best(metric = "roc_auc")
```

Select the best model.
```{r}
best_gb <- gb_fit |>
  select_best(metric = "roc_auc")
best_gb
```

```{r}
# Final workflow
final_wf <- gb_wf |>
  finalize_workflow(best_gb)
final_wf
```

```{r}
# Fit the whole training set, then predict the test cases
final_fit <- 
  final_wf |>
  last_fit(data_split)
final_fit
```

```{r}
# Test metrics
final_fit |> 
  collect_metrics()
```



#### Model Stacking

```{r}
library(stacks)
library(keras)
library(kernlab)

ms_recipe <- 
  recipe(
    los_long ~ ., 
    data = train_data
  ) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())
ms_recipe
```

```{r}
set.seed(203)
folds <- vfold_cv(train_data, v = 5)
```


##### Logistic Regression
```{r}
logit_mod <- 
  logistic_reg(
    penalty = tune(), 
    mixture = tune()
  ) |> 
  set_engine("glmnet", standardize = TRUE)
logit_mod
```

```{r}
logit_wf <- workflow() |>
  add_recipe(ms_recipe) |>
  add_model(logit_mod)
logit_wf
```

```{r}
logit_grid <- grid_regular(
  penalty(range = c(-6, 3)), 
  mixture(),
  levels = c(100, 5)
  )

logit_res <- 
  tune_grid(
    object = logit_wf, 
    resamples = folds, 
    grid = logit_grid,
    control = control_stack_grid()
  )
logit_res
```


##### Random forest
```{r}
rf_mod <- 
  rand_forest(
    mode = "classification",
    # Number of predictors randomly sampled in each split
    mtry = tune(),
    # Number of trees in ensemble
    trees = tune()
  ) |>
  set_engine("ranger")
rf_mod
```

```{r}
rf_wf <- workflow() |>
  add_recipe(ms_recipe) |>
  add_model(rf_mod)
rf_wf
```

```{r}
rf_grid <- grid_regular(
  trees(range = c(100L, 500L)), 
  mtry(range = c(1L, 5L)),
  levels = c(2, 5)
  )

rf_res <- 
  tune_grid(
    object = rf_wf, 
    resamples = folds, 
    grid = rf_grid,
    control = control_stack_grid()
  )
rf_res
```


##### XGBoost
```{r}
gb_mod <- 
  boost_tree(
    mode = "classification",
    trees = 500, 
    tree_depth = tune(),
    learn_rate = tune()
  ) |> 
  set_engine("xgboost")
gb_mod
```

```{r}
gb_wf <- workflow() |>
  add_recipe(ms_recipe) |>
  add_model(gb_mod)
gb_wf
```

```{r}
gb_grid <- grid_regular(
  tree_depth(range = c(3L, 10L)),
  learn_rate(range = c(-3, 2), trans = log10_trans()),
  levels = c(2, 5)
  )

gb_res <- 
  tune_grid(
    object = gb_wf, 
    resamples = folds, 
    grid = gb_grid,
    control = control_stack_grid()
  )
gb_res
```


##### Model stacking

```{r}
mimic_model_st <- 
  # initialize the stack
  stacks() |>
  # add candidate members
  add_candidates(logit_res) |>
  add_candidates(rf_res) |>
  add_candidates(gb_res) |>
  # determine how to combine their predictions
  blend_predictions(
    penalty = 10^(-6:2),
    metrics = c("roc_auc")
    ) |>
  # fit the candidates with nonzero stacking coefficients
  fit_members()
```

```{r}
mimic_model_st
```

```{r}
autoplot(mimic_model_st)
```

```{r}
autoplot(mimic_model_st, type = "members")
```

Top results:
```{r}
autoplot(mimic_model_st, type = "weights")
```

```{r}
collect_parameters(mimic_model_st, "rf_res")
```


##### Final Classification

```{r}
pred <- test_data %>%
  bind_cols(predict(mimic_model_st, ., type = "prob")) %>%
  print(width = Inf)
```

```{r}
yardstick::roc_auc(
  pred,
  truth = los_long,
  contains(".pred_No")
  )
```

```{r}
pred <-
  test_data |>
  select(los_long) |>
  bind_cols(
    predict(
      mimic_model_st,
      test_data,
      type = "class",
      members = TRUE
      )
    ) |>
  print(width = Inf)
```

```{r}
map(
  colnames(pred),
  ~mean(pred$los_long == pull(pred, .x))
  ) |>
  set_names(colnames(pred)) |>
  as_tibble() |>
  pivot_longer(c(everything(), -los_long))
```



### Question 4
Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each machine learning algorithm and the model stacking. Interpret the results. What are the most important features in predicting long ICU stays? How do the models compare in terms of performance and interpretability?

```{r,eval=FALSE}
evaluate_model <- function(model, model_name, test_data) {
  # Predict probabilities
  preds_prob <- predict(model, test_data, type = "prob") |>
    bind_cols(test_data)
  
  # ROC & AUC
  auc <- yardstick::roc_auc(preds_prob, 
                            truth = los_long, 
                            .pred_TRUE) |>
    pull(.estimate)
  
  # Predict class labels
  preds_class <- predict(model, test_data, type = "class") |>
    bind_cols(test_data)
  
  # Compute accuracy
  acc <- yardstick::accuracy(preds_class, 
                             truth = los_long, 
                             estimate = .pred_class) |>
    pull(.estimate)
  
  tibble(Model = model_name, AUC = auc, Accuracy = acc)
}

model_comparison <- bind_rows(
  evaluate_model(final_rf, "Random Forest", test_data),
  evaluate_model(final_gb, "XGBoost", test_data),
  evaluate_model(final_logit, "Logistic Regression", test_data),
  evaluate_model(mimic_model_st, "Stacked Model", test_data)
)

model_comparison
```

```{r}
roc_rf <- final_rf |>
  predict(test_data, type = "prob") |>
  bind_cols(test_data) |>
  roc_curve(truth = los_long, .pred_TRUE) |>
  mutate(Model = "Random Forest")

roc_gb <- final_gb |>
  predict(test_data, type = "prob") |>
  bind_cols(test_data) |>
  roc_curve(truth = los_long, .pred_TRUE) |>
  mutate(Model = "XGBoost")

roc_logit <- final_logit |>
  predict(test_data, type = "prob") |>
  bind_cols(test_data) |>
  roc_curve(truth = los_long, .pred_TRUE) |>
  mutate(Model = "Logistic Regression")

roc_stack <- mimic_model_st |>
  predict(test_data, type = "prob") |>
  bind_cols(test_data) |>
  roc_curve(truth = los_long, .pred_TRUE) |>
  mutate(Model = "Stacked Model")

# Combine ROC data
roc_data <- bind_rows(roc_rf, roc_gb, roc_logit, roc_stack)

ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity, color = Model)) +
  geom_line(size = 1) +
  labs(title = "ROC Curves for Model Comparison", 
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()
```










